Exploring Determinants of Child Mortality Globally (2022): The Role of Vaccination, Healthcare Spending, and Nutrition

### 2024-12-10 paper/paper.qmd
grammar error change

Despite this progress, the pace of reduction has slowed considerably since 2015, decreasing from an annual rate of 3.8% during the Millennium Development Goal (MDG) era (2000–2015) to 2.1% during the Sustainable Development Goal (SDG) era (2015–2022) [@unicef2022underfive]. In 2022 alone, 4.9 million children under the age of five died, equating to approximately 13,400 deaths per day [@unicef2022underfive]. Moreover, the distribution of mortality rates under five remains highly uneven. Developed countries such as Canada and the United States report mortality rates as low as 10 per 1,000 live births, while developing nations—particularly in sub-Saharan Africa—experience rates exceeding 100 per 1,000 live births [@worldbank2024M]. This stark disparity underscores the urgent need for targeted interventions in high-mortality regions.

@@

**Mortality Rate**: The child mortality rate is estimated by the UN Inter-agency Group for Child Mortality Estimation (IGME). The primary data source is each country's vital registration system. In countries such system is lacking, household surveys are conducted and then processed using a statistical model developed collaboratively by UN agencies and academic institutions to derive an estimate for mortality rate [@worldbank2024M].

@@

**Food Production Index**: This index, prepared by the Food and Agriculture Organization (FAO) of the United Nations, measures aggregate food production relative to the 2014–2016 baseline period. It is calculated using price-weighted data on agricultural products for each country yearly. [@worldbank_foodindex2024].

**Health Expenditure per Capita**: Data on health expenditure per capita are produced by the World Health Organization (WHO) following the System of Health Accounts 2011 (SHA 2011) framework. This framework tracks all health-related spending in a country over a defined time period [@worldbank_health_expenditure_us].

@@

The response variable in this study is the child mortality rate in each country in year 2022, measured as the number of deaths of children under five years old per 1,000 live births in year 2022.

@@

```{r}
#| label: fig-mor
#| fig-cap: Data Analysis for Mortality Rate for children under five of different countries. The histogram on the top left shows that the original distribution is skewed to the right. The histogram on the top right is the distribution after log transformation and is approximately normal.
#| fig-subcap: ["Dsitribution of Mortality rate before log tranformation", "Distribution of Mortality rate after log transformation"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

**Food Production Index:** 
The Food Production Index measures the relative level of agricultural production for each country in year 2022, using the 2014–2016 baseline as a reference point [@worldbank]. This index is price-weighted and reflects agricultural productivity across countries. As illustrated in @fig-food-1, the index ranges from 60 to 180, with a mean of around 110 and an approximately normal distribution. Scatter plot between food production index and log of child mortality in @fig-food-2 shows a slightly positive linear relationship.This relationship is examined in greater detail in @sec-results.
```{r}
#| label: fig-food
#| fig-cap: Data Analysis of Predictor Food Production Index. The histogram on the left shows a approximately normal distrubution. The scatterplot on the right indicates there is a slight positive relationship between food production index and mortality rate
#| fig-subcap: ["Counts of Food production index for each country, measured as a basis of 2014-1016", "Relationship of each countries' Food production index VS. Log of Mortality rate"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

ggplot(analysis_data, aes(x = Food, y = Mortality)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkred", fill = "pink", se = TRUE, alpha = 0.3) +
  labs(
    x = "Food Production Index",
    y = "Mortality Rate",
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12)
  )
```
**Current Health Expenditure:**
The Current Health Expenditure per Capita captures health spending in US dollars for each individual in year 2021, calculated under the System of Health Accounts (SHA 2011) framework. The raw data range from $0 to $12,000 and exhibit extreme right skewness, as shown in @fig-health-1. To address this, a log transformation was applied, resulting in a normalized range from 0 to 10, with a central tendency near 6 (@fig-health-2). Analysis of the relationship between log-transformed health expenditure and log mortality rate (@fig-health-3) demonstrates a significant negative linear trend, indicating that higher health expenditures are associated with lower mortality rates. This relationship is examined in greater detail in @sec-results.
```{r}
#| label: fig-health
#| fig-cap: Data Analysis of health expenses per capita. The histogram on the top left shows that the original distribution is skewed to the right. The histogram on the top right is the distribution after log transformation and is approximately normal.The scatterplot on the bottom left indicates there is a negative relationship between log of health expenses and Mortality rate.
#| fig-subcap: ["Counts of each countries' Health Expenditure Per Capita, Measured in US$. Before log transformation", "Counts of each countries' Health Expenditure Per Capita, Measured in US$. After log transformation", "Relationship of each countries' Health Expenditure Per Capita, Measured in US$ VS. Log of Mortality"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

ggplot(analysis_data, aes(x = Health_expense, y = Mortality)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkred", fill = "pink", se = TRUE, alpha = 0.3) +
  labs(
    x = "Log of Health Expense per capita",
    y = "Mortality Rate",
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12)
  )
```
**DPT Vaccine Coverage:**
DPT vaccine coverage rate reflects the percentage of children aged 12–23 months who received vaccinations against diphtheria, pertussis, and tetanus in year 2022. Coverage ranges from 30% to 100%, with a left-skewed distribution, as depicted in @fig-DPT-1. Analysis of the association between vaccine coverage and mortality rates (@fig-DPT-2) shows a strong negative relationship, suggesting that increased vaccine coverage is correlated with lower mortality rates. This finding will be explored further in @sec-model.
```{r}
#| label: fig-DPT
#| fig-cap: Data Analysis of DPT vaccine coverage. The histogram on the left shows that the distribution DPT vaccine coverage of is skewed to the left. The scatterplot on the right indicates there is a negative relationship between DPT vaccine coverage and Mortality rate.
#| fig-subcap: ["Counts of each countries' DPT vaccine percentage", "Relationship of each countries' DPT Vaccine Percentage VS. Log of Mortality"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

ggplot(analysis_data, aes(x = Vacinne, y = Mortality)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkred", fill = "pink", se = TRUE, alpha = 0.3) +
  labs(
    x = "DPT vaccine percentage",
    y = "Mortality Rate",
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12)
  )
```
## Missing Data and Time Inconsistancy {#sec-data-missing}
The health expenditure variable was collected in 2021, while other variables reflect 2022 data. This discrepancy is due to the unavailability of more recent health expenditure data. These 2021 values are assumed to approximate 2022 conditions sufficiently for analysis.

Additionally, some countries' data are not collected and were excluded from the analysis. The implications of these omissions and time inconsistancy are addressed in @sec-lim.

@@

Let $y_i$ be the logarithm of under-five mortality rate for nation $i$. We define the following predictors:

@@

@fig-stanareyouokay-1 displays the trace plot check, where no irregularities or concerning patterns are detected. Finally, the R-hat plot (@fig-stanareyouokay-2) shows all values close to one, confirming the convergence of the Markov chains. Detailed information could be foun in @sec-model-details.

@@

Our model results are summarized in @tbl-modelresults .We will discuss them one by one

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory model of Mortality rate for each country based on food production index, DPT vaccine coverage, and health expense"

@@

## Overview
The United Nations’ Sustainable Development Goal (SDG) 3.2 aims to reduce the mortality rate of children under five to fewer than 25 per 1,000 live births by 2030 [@who2024]. While substantial progress has been achieved globally, meeting this ambitious target within the remaining timeline of six years presents significant challenges. Although global trends indicate a decline in child mortality rates, the pace of this reduction has slowed since 2015 [@who2024].

@@

Our study shows that reducing child mortality rate requires a dual focus on expanding vaccination coverage and optimizing healthcare spending. Vaccination programs, particularly for diseases like diphtheria, pertussis, and tetanus, represent a direct strategy for reducing child mortality. Simultaneously, enhancing public healthcare spending can address systemic issues such as inadequate maternal care and limited access to trained professionals.

Oh the other hand, efforts must be tailored to the specific needs and capacities of individual countries. International organizations can play a pivotal role by providing financial support, technical expertise, and policy guidance to countries with high child mortality rates. By prioritizing interventions with clear and measurable impacts, the global community can move closer to the goal of ensuring that no child dies from preventable causes.

## Limitation and next steps {#sec-lim}
This study has several limitations that warrant consideration in future research. First, as discussed in @sec-data-missing, data from several countries is unavailable. While these nations typically have smaller populations, their exclusion may introduce biases and raises ethical concerns by neglecting their unique challenges. Second, the most recent data used in this study is from 2022, and there are temporal inconsistencies among the predictors. This is particularly important because both the predictors and the response variable are time-sensitive. Access to more current and synchronized datasets would enable a more accurate and relevant analysis, reflecting recent developments in healthcare and child mortality trends.

>>>>

Despite this progress, the pace of reduction has slowed considerably since 2015, decreasing from an annual rate of 3.8% during the Millennium Development Goal (MDG) era (2000–2015) to 2.1% during the Sustainable Development Goal (SDG) era (2015–2022) [@unicef2022underfive]. In 2022 alone, 4.9 million children under the age of five died, equating to approximately 13,400 deaths per day [@unicef2022underfive]. Moreover, the distribution of mortality rates under five remains highly uneven. Developed countries such as Canada and the United States report mortality rates as low as 10 per 1,000 live births while developing nations—particularly in sub-Saharan Africa—experience rates exceeding 100 per 1,000 live births [@worldbank2024M]. This stark disparity underscores the urgent need for targeted interventions in high-mortality regions.

@@

**Mortality Rate**: The child mortality rate is estimated by the UN Inter-agency Group for Child Mortality Estimation (IGME). The primary data source is each country's vital registration system. In countries where such a system is lacking, household surveys are conducted and then processed using a statistical model developed collaboratively by UN agencies and academic institutions to derive an estimate of the mortality rate [@worldbank2024M].

@@

**Food Production Index**: This index, prepared by the Food and Agriculture Organization (FAO) of the United Nations, measures aggregate food production relative to the 2014–2016 baseline period. It is calculated yearly using price-weighted data on agricultural products for each country. [@worldbank_foodindex2024].

**Health Expenditure per Capita**: Data on health expenditure per capita are produced by the World Health Organization (WHO) following the System of Health Accounts 2011 (SHA 2011) framework. This framework tracks all health-related spending in a country over a defined period [@worldbank_health_expenditure_us].

@@

The response variable in this study is the child mortality rate in each country in 2022, measured as the number of deaths of children under five years old per 1,000 live births in 2022.

@@

```{r}
#| label: fig-mor
#| fig-cap: Data Analysis for Mortality Rate for children under five in different countries. The histogram on the top left shows that the original distribution is skewed to the right. The histogram on the top right is the distribution after log transformation and is approximately normal.
#| fig-subcap: ["Distribution of Mortality rate before log transformation", "Distribution of Mortality rate after log transformation"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

**Food Production Index:** 
The Food Production Index measures the relative level of agricultural production for each country in 2022, using the 2014–2016 baseline as a reference point [@worldbank]. This index is price-weighted and reflects agricultural productivity across countries. As illustrated in @fig-food-1, the index ranges from 60 to 180, with a mean of around 110 and an approximately normal distribution. The scatter plot between the food production index and log of child mortality in @fig-food-2 shows a slightly positive linear relationship. This relationship is examined in greater detail in @sec-results.
```{r}
#| label: fig-food
#| fig-cap: Data Analysis of Predictor Food Production Index. The histogram on the left shows an approximately normal distribution. The scatterplot on the right indicates there is a slight positive relationship between food production index and mortality rate.
#| fig-subcap: ["Counts of Food production index for each country, measured as a basis of 2014-1016", "Relationship of each countries' Food production index VS. Log of Mortality rate"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

ggplot(analysis_data, aes(x = Food, y = Mortality)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkred", fill = "pink", se = TRUE, alpha = 0.3) +
  labs(
    x = "Food Production Index",
    y = "Mortality Rate",
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12)
  )
```
**Current Health Expenditure:**
The Current Health Expenditure per Capita captures health spending in US dollars for each individual in 2021, calculated under the System of Health Accounts (SHA 2011) framework. The raw data range from $0 to $12,000 and exhibit extreme right skewness, as shown in @fig-health-1. To address this, a log transformation was applied, resulting in a normalized range from 0 to 10, with a central tendency near 6 (@fig-health-2). Analysis of the relationship between log-transformed health expenditure and log mortality rate (@fig-health-3) demonstrates a significant negative linear trend, indicating that higher health expenditures are associated with lower mortality rates. This relationship is examined in greater detail in @sec-results.
```{r}
#| label: fig-health
#| fig-cap: Data Analysis of health expenses per capita. The histogram on the top left shows that the original distribution is skewed to the right. The histogram on the top right is the distribution after log transformation and is approximately normal.The scatterplot on the bottom left indicates there is a negative relationship between log of health expenses and Mortality rate.
#| fig-subcap: ["Counts of Health Expenditure Per Capita for each country (measured in US$) before log transformation", "Counts of each country's Health Expenditure Per Capita, (measured in US$) after log transformation", "Relationship of each countries' Health Expenditure Per Capita, Measured in US$ VS. Log of Mortality"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

ggplot(analysis_data, aes(x = Health_expense, y = Mortality)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkred", fill = "pink", se = TRUE, alpha = 0.3) +
  labs(
    x = "Log of Health Expense per capita",
    y = "Mortality Rate",
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12)
  )
```
**DPT Vaccine Coverage:**
DPT vaccine coverage rate reflects the percentage of children aged 12–23 months who received vaccinations against diphtheria, pertussis, and tetanus in 2022. Coverage ranges from 30% to 100%, with a left-skewed distribution, as depicted in @fig-DPT-1. Analysis of the association between vaccine coverage and mortality rates (@fig-DPT-2) shows a strong negative relationship, suggesting that increased vaccine coverage is correlated with lower mortality rates. This finding will be explored further in @sec-model.
```{r}
#| label: fig-DPT
#| fig-cap: Data Analysis of DPT Vaccine Coverage. The histogram on the left shows that the distribution of DPT vaccine coverage is skewed to the left. The scatterplot on the right indicates there is a negative relationship between DPT vaccine coverage and Mortality rate.
#| fig-subcap: ["Counts of each country's DPT vaccine percentage", "Relationship of each country's DPT Vaccine Percentage VS. Log of Mortality"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

ggplot(analysis_data, aes(x = Vacinne, y = Mortality)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkred", fill = "pink", se = TRUE, alpha = 0.3) +
  labs(
    x = "DPT vaccine percentage",
    y = "Mortality Rate",
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12)
  )
```
## Missing Data and Time Inconsistency {#sec-data-missing}
The health expenditure variable was collected in 2021, while other variables reflect 2022 data. This discrepancy is due to the unavailability of more recent health expenditure data. These 2021 values are assumed to approximate 2022 conditions sufficiently for analysis.

Additionally, some countries' data were not collected and were excluded from the analysis. The implications of these omissions and time inconsistency are addressed in @sec-lim.

@@

Let $y_i$ be the logarithm of the under-five mortality rate for nation $i$. We define the following predictors:

@@

@fig-stanareyouokay-1 displays the trace plot check, where no irregularities or concerning patterns are detected. Finally, the R-hat plot (@fig-stanareyouokay-2) shows all values close to one, confirming the convergence of the Markov chains. Detailed information could be found in @sec-model-details.

@@

Our model results are summarized in @tbl-modelresults. We will discuss them one by one.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory model of Mortality rate for each country based on food production index, DPT vaccine coverage, and health expense."

@@

## Overview
The United Nations’ Sustainable Development Goal (SDG) 3.2 aims to reduce the mortality rate of children under five to fewer than 25 per 1,000 live births by 2030 [@who2024]. While substantial progress has been achieved globally, meeting this ambitious target within the remaining six years presents significant challenges. Although global trends indicate a decline in child mortality rates, the pace of this reduction has slowed since 2015 [@who2024].

@@

Our study shows that reducing the child mortality rate requires a dual focus on expanding vaccination coverage and optimizing healthcare spending. Vaccination programs, particularly for diseases like diphtheria, pertussis, and tetanus, represent a direct strategy for reducing child mortality. Simultaneously, enhancing public healthcare spending can address systemic issues such as inadequate maternal care and limited access to trained professionals.

On the other hand, efforts must be tailored to the specific needs and capacities of individual countries. International organizations can play a pivotal role by providing financial support, technical expertise, and policy guidance to countries with high child mortality rates. By prioritizing interventions with clear and measurable impacts, the global community can move closer to the goal of ensuring that no child dies from preventable causes.

## Limitation and next steps {#sec-lim}
This study has several limitations that warrant consideration in future research. First, as discussed in @sec-data-missing, data from several countries is unavailable. While these nations typically have smaller populations, their exclusion may introduce biases and raise ethical concerns by neglecting their unique challenges. Second, the most recent data used in this study is from 2022, and there are temporal inconsistencies among the predictors. This is particularly important because both the predictors and the response variable are time-sensitive. Access to more current and synchronized datasets would enable a more accurate and relevant analysis, reflecting recent developments in healthcare and child mortality trends.
